Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-05-08 paper.tex
Something incomprehensible about rotations

Our implementation fixes $x_1$ and $x_2$ to be parallel to one of the axes $X$,
$Y$ or $Z$ with restriction $x_1 \perp x_2$ (\cref{fig:pattern})
\cite{dimitrakopoulos2010high}. Thus, the output has dimensions $D_1 \times D_2$
where $D_1$ and $D_2$ are dimensions of the input array along selected
axes. Note that shift operators with such $x_1$ and $x_2$ represent a 3-point
function computed within a 3D domain, but along a predefined direction (see
below).
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.5\linewidth]{images/pattern.png}
  \caption[]{The pattern used for calculation of $S_3$ function in our
    implementation.}
  \label{fig:pattern}
\end{figure}

@@

@@

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix
$\Sigma = \left\{ \sigma_{ij} \right\}$ for coordinates of points in the phase of
interest:
\begin{equation}
  \sigma_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $\Sigma$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample makes the direction of
anisotropy parallel to one of the axes, so anisotropy can be better represented in
the correlation function \cref{fig:aniso}. In our implementation we use rotation
with nearest-neighbor interpolation which affects quality of computations as
discussed in \cref{sec:rotqua}.
\begin{figure*}[!htp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.3\linewidth, frame]{images/aniso.png} &
    \includegraphics[width=0.3\linewidth, frame]{images/aniso-rot.png} & \\
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png} \\
  \end{tabular}
  \caption{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is elongated along the axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to the algorithm described in
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

>>>>

@@

\subsection{Open-source implementation}
The described algorithms are implemented in our \code{CorrelationFunctions.jl}
package written in Julia and have the same interface:
\begin{verbatim}
corrfn(input, phase, ps1, ps2)
\end{verbatim}
where \code{corrfn} stands for \code{s3} ($S_3$), \code{c3} ($C_3$),
\code{surf3} ($F_{sss}$) etc., \code{input} is the input array, \code{phase} is
a phase of interest and \code{ps1} and \code{ps2} are arrays of points in which
correlation function is calculated. Firstly, \code{corrfn} does all necessary
preparations for the input array (e.g. edge extraction or connected component
labeling) and then calculates the correlation function for each pair of
arguments from \code{ps1} and \code{ps2}.

Our implementation provides a predefined right triangle pattern where arguments
of a correlation function $x_1$ and $x_2$ are fixed to be parallel to one of the
axes $X$, $Y$ or $Z$ with restriction $x_1 \perp x_2$ (\cref{fig:pattern})
\cite{dimitrakopoulos2010high}. The output has dimensions $D_1 \times D_2$ where
$D_1$ and $D_2$ are dimensions of the input array along selected axes. Note that
shift operators with such $x_1$ and $x_2$ represent a 3-point function computed
within a 3D domain, but along a predefined direction (see below).
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.5\linewidth]{images/pattern.png}
  \caption[]{The pattern used for calculation of $S_3$ function in our
    implementation.}
  \label{fig:pattern}
\end{figure}

If it is desired, a user can use a custom pattern more appropriate to the
situation.

@@

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix
$\Sigma = \left\{ \sigma_{ij} \right\}$ for coordinates of points in the phase of
interest:
\begin{equation}
  \sigma_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $\Sigma$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample (or its inverse to the
pattern) makes the direction of anisotropy coaligned with the pattern, so
anisotropy can be better represented in the correlation function. For the sake
of visual clarity in this study we choose to rotate the image instead of the
pattern \cref{fig:aniso}. This has such consequences as reduced accuracy of
computed functions due to necessity to interpolate the image during
rotation. Although we show in \cref{sec:rotqua} that even nearest-neighbor
interpolation results only in insignificant loss of quality. Rotation of the
pattern has same drawbacks related to interpolation of the sample which are most
noticeable at small correlation lengths. The both methods (rotation of the
sample or the pattern) produce nearly equivalent result if the sample is
homogeneous and an average pore size is much less than the sample's size (see
\cref{sec:rotrot})
\begin{figure*}[!htp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.3\linewidth, frame]{images/aniso.png} &
    \includegraphics[width=0.3\linewidth, frame]{images/aniso-rot.png} & \\
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png} \\
  \end{tabular}
  \caption{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is elongated along the axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to the algorithm described in
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

\section{Difference between rotating the input image or the sampling pattern}
\label{sec:rotrot}
Let's take the noise from \cref{sec:rotqua} with lesser sizes of pores and
calculate $S_3$ function for it, preliminarily rotating the noise around Z-axis
counter-clockwise using nearest-neighbor interpolation (the first case) or
rotating the pattern around the same axis clock-wise (the second case). Then we
compute a difference between the results point-wise using a formula
\begin{equation}
  \varepsilon = 2\frac{|S_3^{(1)} - S_3^{(2)}|}{S_3^{(1)} + S_3^{(2)}}
  \label{eq:reldiff}
\end{equation}
The average of point-wise differences is at maximum for the angle of rotation
$\theta = \frac{\pi}{4}$ and equals to 0.8\%. The maximal difference is about
10\% for very small correlation length (about 2-3 pixels). These two methods of
rotating give minimal difference if applied to a homogeneous medium with an
average pore size much less than a size of the input (\cref{fig:rotdiff}).
\begin{figure*}[tp]
  \centering
  \subfigure[A sample of value noise with small pores (compared to a size of the
    sample, $300 \times 300 \times 300$)]{
    \includegraphics[width=0.4\linewidth]{images/noise2.png}
    \label{fig:rotdiff-noise}}
  \hfill
  \subfigure[Difference between two methods of rotation]{
    \includegraphics[width=0.4\linewidth]{images/rotdiff.png}
    \label{fig:rotdiff-error}}
  \caption[]{A plot depicting difference between two rotation methods: rotation
    of the sample and of the pattern. Dependency of an average point-wise
    difference \cref{eq:reldiff} from a rotation angle is shown. The rotation is
    performed around $Z$ axis counter-clockwise (sample) and clockwise (pattern).}
  \label{fig:rotdiff}
\end{figure*}
